---
title: "Visualizing 2D Categorical and Continuous Variables"
output: 
  html_document:
    toc: TRUE
    toc_float:
      collapsed: FALSE
    theme: yeti
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "..") }) 
date: "2023-06-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", message = FALSE, warning = FALSE)
library(tidyverse)
```

```{r}
ohtani_batted_balls <- read_csv("https://shorturl.at/mnwL1")
head(ohtani_batted_balls)
```
```{r}
ohtani_batted_balls <- ohtani_batted_balls %>% 
  filter(pitch_type != "null") %>% 
  mutate(pitch_type = fct_recode(pitch_type,
"Changeup" = "CH", "Breaking ball" = "CU", "Fastball" = "FC","Fastball" = "FF","Fastball" = "FS","Breaking ball" = "KC", "Fastball" = "SI", "Breaking ball" = "SL"))

table(ohtani_batted_balls$pitch_type)
```

## Chi-Squared Distribution

**Question:** Are all pitch types equally likely to occur?

To answer this, we can preform a chi-squared test!

**Hypotheses:**

* $H_0: p_1 = p_2 = ... = p_k$

* $H_a:$ at least two of p_a for i = 1, 2,..,k are not equal to one another.

```{r}
chisq.test(table(ohtani_batted_balls$pitch_type))
```

**Conclusion:** At a significance level of 0.05, we reject the null hypothesis in favor of there being very strong evidence (p-value approximately 0) that all pitch types are *not* equally likely to occur.

## 2D Categorical Visualization 

```{r}
ohtani_batted_balls %>%
ggplot(aes(x = batted_ball_type,
fill = pitch_type)) +
  geom_bar() + 
  theme_bw()
```

```{r}
ohtani_batted_balls %>%
ggplot(aes(x = batted_ball_type,
fill = pitch_type)) +
geom_bar(position = "dodge") +
  theme_bw()
```


```{r}
table("Pitch type" = ohtani_batted_balls$pitch_type,
"Batted ball type" = ohtani_batted_balls$batted_ball_type)

proportions(table(ohtani_batted_balls$pitch_type, ohtani_batted_balls$batted_ball_type))
```

```{r}
## joint probabilities table via dplyr

library(gt)
ohtani_batted_balls %>%
group_by(batted_ball_type, pitch_type) %>%
summarize(joint_prob = n() / nrow(ohtani_batted_balls)) %>%
pivot_wider(names_from = batted_ball_type, values_from = joint_prob,
values_fill = 0) %>%
gt()
```

### Visualizing Independence between 2 Categorical Variables

```{r}
mosaicplot(table(ohtani_batted_balls$pitch_type, ohtani_batted_balls$batted_ball_type),
           shade = TRUE, #adds pearson residuals to mosiac plot
           # rij approx 0 means observed counts are close to expected counts
           # if abs(rij) > 2 means "signficant" at alpha = 0.05
           main = "Relationship between batted ball and pitch type")

```

## Continuous by Categorical

```{r}
ohtani_batted_balls %>%
ggplot(aes(x = pitch_type,
y = exit_velocity)) +
geom_violin() +
geom_boxplot(width = .2)+
  theme_bw()
```

```{r}
ohtani_batted_balls %>%
ggplot(aes(x = exit_velocity,
color = pitch_type)) +
stat_ecdf() +
theme_bw() +
theme(legend.position = "bottom")
```

```{r}
ohtani_batted_balls %>%
ggplot(aes(x = exit_velocity,
fill = pitch_type)) +
geom_histogram(alpha = .25, position = "identity") +
theme_bw() +
  theme(legend.position = "bottom")
```

```{r}
ohtani_batted_balls %>%
ggplot(aes(x = exit_velocity)) +
geom_histogram(aes(fill = pitch_type)) +
theme_bw() +
facet_wrap(~ pitch_type, ncol = 2)
```

```{r}
ohtani_batted_balls %>%
ggplot(aes(x = exit_velocity)) +
geom_histogram(aes(fill = pitch_type)) +
theme_bw() +
facet_grid(pitch_type ~., margins = TRUE)
```

### Facets make it easy to move beyond 2D

```{r}
ohtani_batted_balls %>%
ggplot(aes(x = pitch_type,
fill = batted_ball_type)) +
geom_bar() + theme_bw() +
facet_wrap(~ outcome, ncol = 5) +
theme(legend.position = "bottom", axis.text = element_text(size = 4))
```

## 2D Continuous Relationships 

```{r}
ohtani_batted_balls %>%
ggplot(aes(x = exit_velocity,
y = launch_angle)) +
geom_point(aes(color = batted_ball_type)) +
theme_bw()
```


