---
title: "Hierarchical Clustering"
output: 
  html_document:
    toc: TRUE
    toc_float:
      collapsed: FALSE
    theme: yeti
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "..") })
date: "2023-06-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", message = FALSE, warning = FALSE)
library(tidyverse)
library(dslabs)
library(seriation)
library(ggdendro)
library(protoclust)
```

----

# Data Set-up

```{r}
gapminder <- as_tibble(gapminder)
head(gapminder)
```

```{r}
clean_gapminder <- gapminder %>%  filter(year == 2011, !is.na(gdp)) %>%  mutate(log_gdp = log(gdp))
clean_gapminder
```

## Standardization

```{r}
clean_gapminder <- clean_gapminder %>%
  mutate(s_log_gdp = as.numeric(scale(log_gdp, center = TRUE, scale = TRUE)), s_life_exp = as.numeric(scale(life_expectancy,        center = TRUE, scale = TRUE))) 

clean_gapminder %>%
  ggplot(aes(x = s_log_gdp, y = s_life_exp))+
  geom_point() +
  theme_bw() +
  coord_fixed()
```

# Computing the distance matrix

**Pairwise Euclidean Distance:**

```{r}
gap_dist <- dist(dplyr::select(clean_gapminder, s_log_gdp, s_life_exp))
```

**Crafting the Matrix:**

```{r}
gap_dist_matrix <- as.matrix(gap_dist)
rownames(gap_dist_matrix) <- clean_gapminder$country
colnames(gap_dist_matrix) <- clean_gapminder$country
head(gap_dist_matrix[1:3, 1:3])
```
# Plotting Similarities 

```{r}

long_dist_matrix <- as_tibble(gap_dist_matrix) %>%
  mutate(country1 = rownames(gap_dist_matrix)) %>%
  pivot_longer(cols = -country1, names_to = "country2", values_to = "distance")
         
long_dist_matrix %>%
  ggplot(aes(x = country1, y = country2, fill = distance)) +
  geom_tile() +
  theme_bw() +
  theme(axis.text = element_blank(), axis.ticks = element_blank(), legend.position = "bottom") +  scale_fill_gradient(low = "darkorange", high = "darkblue")
```



```{r}
gap_dist_seriate <- seriate(gap_dist)

gap_order <- get_order(gap_dist_seriate)

gap_countries_order <-
  as.character(clean_gapminder$country[gap_order]) 
                                       long_dist_matrix$country1 <- as_factor(long_dist_matrix$country1) 
                                       long_dist_matrix$country2 <- as_factor(long_dist_matrix$country2) 
                                       long_dist_matrix %>%                                         mutate(country1 = fct_relevel(country1,            gap_countries_order), country2 = fct_relevel(country2,                     gap_countries_order)) %>%
                                         ggplot(aes(x = country1, y = country2,              fill = distance)) +  geom_tile() + theme_bw() +  theme(axis.text = element_blank(),         axis.ticks = element_blank(), legend.position = "bottom") +
                                         scale_fill_gradient(low = "darkorange", high = "darkblue")
```
# Agglomerative Hierarchical Clustering

Pretend all *n* observations are their own cluster

* Step 1: Compute the pairwise dissimilarities between each cluster (e.g., distance matrix)

* Step 2: Idenitfy the pair of clusters that are least dissimilar

* Step 3: Fuse these two clusters into a new cluster

* Repeat Steps 1 to 3 until all observations are in the same cluster

* **Bottom-up** agglomerative clusters that forms a tree/hierarchy of merging 

## How do we Define Dissimilarity between Clusters?

**We need a linkage function!**

* Complete linkage: uses maximum value of these dissimilarities (i.e., distance)

* Single linkage: uses minimum value of these dissimilarities (i.e., distance)

* Average Linkage: uses average value of these dissimilarities (i.e., distance)

**Define dissimilarity between two clusters based on our initial dissimilarity matrix between observations**

## Complete Linkage Example 

```{r}
gap_complete_hclust <- hclust(gap_dist, method = "complete")
```

```{r}
clean_gapminder %>%  mutate(country_clusters =         as.factor(cutree(gap_complete_hclust, k = 4))) %>%
  ggplot(aes(x = s_log_gdp, y = s_life_exp,
             color = country_clusters)) +  geom_point() +
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom")
```

### Dendrogram

```{r}
ggdendrogram(gap_complete_hclust,             theme_dendro = FALSE, labels = FALSE,             leaf_labels = FALSE) +
  labs(y = "Dissimilarity between clusters")  +
  theme_bw() +
  theme(axis.text.x = element_blank(),         axis.title.x = element_blank(), axis.ticks.x = element_blank(), panel.grid = element_blank())
```

* Each leaf = one observation

* Height of branch represents the dissimilarity between clusters (Horizontal position along the x-axis means nothing after the first step)

**You can specify the height to cut with h (height) instead of k**

```{r}
clean_gapminder %>%  mutate(country_clusters =         as.factor(cutree(gap_complete_hclust, h = 5))) %>%
  ggplot(aes(x = s_log_gdp, y = s_life_exp,
             color = country_clusters)) +  geom_point() +
  ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom")
```

**NOTE: YOU WILL GET DIFFERENT RESULTS BASED ON HOW YOU DEFINE THE LINKAGE FUNCTION**

## More Linkage Functions

* Centroid Linkage: Computes the dissimilarity between the centroid for cluster 1 and the centroid for cluster 2 (i.e., the distance between the averages of the two clusters)

* Ward's linkage: Merges a pair of clusters to minimize the within-cluster variance (i.e., aim is to minimize the objective function from *K-means*)

* Minimax Linkage

Each cluster is defined **by a prototype** observation (most representative)

**Identify the point whose farthest point is closest**

Use this minimum-maximum distance as the measure of cluster dissimilarity 

Dendrogram interpretation: each point is less than or equal to h in dissimilarity to the the prototype of the cluster 

* **Cluster centers are chosen among the observations themselves -- hence the prototype**

## Minimax Linkage Example 

**Dendrogram**

```{r}
gap_minimax <- protoclust(gap_dist)
plot(gap_minimax)
```

**Scatterplot**

```{r}
minimax_country_clusters <-   protocut(gap_minimax, k = 4)

clean_gapminder %>%
  mutate(country_clusters =     as.factor(minimax_country_clusters$cl)) %>%
  ggplot(aes(x = s_log_gdp, y = s_life_exp,             color = country_clusters)) +  geom_point() +   ggthemes::scale_color_colorblind() +
  theme_bw() +
  theme(legend.position = "bottom")
```

**To find prototypes:**

```{r}
minimax_country_clusters$protos
```

Indices of the prototypes (in the order of the clusters)

**Finding countries with these indices:**

```{r}
clean_gapminder %>%  dplyr::select(country, gdp, life_expectancy,                 population, infant_mortality) %>%  slice(minimax_country_clusters$protos)
```

**How are these clusters related to the continents?**

```{r}
table("Clusters" = minimax_country_clusters$cl, "Continents" = clean_gapminder$continent)
```


